Exploiting First-Class Arrays in Fortran for Accelerator Programming

### 2010-12-14 opencl/papers/source-to-source.tex
Rough draft 0 stage

@@

This sections show short Fortran code examples and OpenCL equivalent.  The
notation uses uppercase for arrays and lowercase for scalar quantities.
Interior and region array copies are denoted by an i or an r preceeding the array.
For example, {\tt iH = interior(H, halo)} is a local copy of interior region
of array {\tt H} (representing height) in the shallow water code.

@@

@@

\begin{verbatim}
   V = V - iA*(V - v_rest)                  !! Fortran

   V[l] = V[l] - iA[lex]*(V[l] - v_rest);   // OpenCL equivalent
\end{verbatim}

\subsubsection{where construct}

A neuron in PetaVision fires (activity is set to 1)
whenever the action potential is greater than some threshold.
This is easily expressed with a where construct.

\begin{verbatim}
   where (V > Vth)                        !! Fortran
      iA = 1
   elsewhere
      iA = 0
   end where

   iA[lex] = (V[l] > Vth[l]) ? 1 : 0;     // OpenCL equivalent
\end{verbatim}

\subsection{New functions}

%\subsubsection{transfer_halo}
\subsubsection{region}

One of the state variables in the shallow water code is H, effectively the
height of the water.  This variable has a halo (ghost-cell region) surrounding
the interior of the grid to handle boundary conditions.  When using
MPI, the halo regions contains values from adjacent processors that must be
updated with new values at each time step.  This is accomplished with the
{\tt transfer\_halo} function.  The shallow water code assumes a five point stencil
so the state variables are extended by 2 in each dimension (e.g., by one to the
left, right, up, and down).

\begin{verbatim}
halo = [1,1,1,1]
iH => region(H, halo, transfer=false)
\end{verbatim}

>>>>

\begin{figure}[!t]
\centering
\includegraphics[width=2.5in]{cl-memory.pdf}
\caption{A schematic of global memory for an array and its copy stored in local memory
for four thread groups.}
\label{fig:cl-memory}
\end{figure}

@@

This sections outlines the OpenCL equivalent syntax for portions of the
Fortran shallow-water code described in the previous section.  The notation
uses uppercase for arrays and lowercase for scalar quantities.  Interior and
region array copies are denoted by an i or an r preceeding the array.  For
example, {\tt iH = interior(H, halo)} is a local copy of interior region of
array {\tt H} (representing height) in the shallow water code.

@@

\subsubsection{function and variable declarations}

Fortran kernel procedures have direct correspondence with OpenCL equivalents.
For example, {\tt wave_advance} is transformed as {\tt __kernel void
  wave_advance(__ global float * H, ..., float dt);}.  The global state
variables have local equivalents, e.g., {\tt __local float
  H_local[MAX_LOCAL_SIZE]}, declared with the appropriate size and are copied
to local memory by the compiler with inlined library functions.  The flux
array temporaries (region variables) are declared similarly but initialized to
zero with inlined library functions.  Interior variables are simple scalars,
e.g., {\tt float iH;}.  Region variables cannot be scalar objects because
regions are shifted and thus \emph{shared} by threads within a thread group.

@@

Array syntax transforms nearly directly to OpenCL code.  For example, interior
variables are particularly straightforward as they as scalar quantities in
OpenCL,
\begin{verbatim}
   iH = iH + (dt/dx) * ( region(Ux, face_lt) - region(Ux, face_rt) )
           + (dt/dy) * ( region(Vy, face_dn) - region(Vy, face_up) );
\end{verbatim}

Region variables are more complicated because they are arrays.
\begin{verbatim}
   Hx[l] = 0.5 * ( region(H_local, face_lt) + ... );
\end{verbatim}
where {\tt l = LX + LY*(NLX+halo(0)+halo(1)))} is a local index variable, {\tt
  LX = get_local_id(0)} is the local thread id in the $x$ dimension, {\tt LY =
  get_local_id(1)} is the local thread id in the $y$ dimension, {\tt NLX =
  get_local_size(0)} is the size of the thread group in the $x$ dimension, and
the {\tt get_local_id} and {\tt get_local_size} functions are defined by the
OpenCL compiler.

\subsection{Performance measurements}

Performance measurements were made comparing the transformed code with
different versions of the serial shallow-water code.  The serial versions
include the original written in C and two Fortran versions: one using copy
semantics for the regions and one using reference semantics.  We also compare
with a hand written OpenCL implementation that was optimized for local memory
usage (no array temporaries).  The measurments were made using an NVIDIA Fermi
GPU (TODO - need model numbers) running at XXX Hz and a VENDOR hexacore CPU
running at XXX Hz.  The compilers were gfortran and gcc VERSIONS with
optimization of -O3.

The performance comparison is shown in Figure~\ref{fig:cl-performance} for
varying array sizes (of the state variables).  The time represents an average
of 100 iterations of the outer time-advance loop calling the OpenCL kernel.
This tight loop keeps the OpenCL kernel supplied with threads to take
advantage of potential latency hiding by the NVIDIA GPU.  Any serial code
within this loop would reduce the measured values.

As can be seen in Figure~\ref{fig:cl-performance}, the transformed get very
good results of roughly 65 times speedup over the best Fortran code and YYY
speedup over the original C code.  It comes with XXX \% of the hand coded
OpenCL version.

\begin{figure}[!t]
\centering
\includegraphics[width=2.5in]{cl-performance.pdf}
\caption{Performance comparison for varying array size.}
\label{fig:cl-performance}
\end{figure}
